---
title: |
  <center> S-052: Intermediate and Advanced Statistical Methods </center>
  <center> for Applied Educational Research </center>
  <center> Spring 2026 </center>
  <center> Week 3 Data Analytic Memo Starter Code </center>
author: "Sara Ji and Bailey Buchanan"
date: "Last Modified: Feb 2026"
output: pdf_document
---

### RQ: What is the causal effect of technology-led afterschool on math scores?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load in packages
```{r packages, warning = FALSE, message = FALSE}
# These packages will be helpful for your analysis
library(foreign)    # For reading various data formats
library(haven)      # For reading Stata files
library(tidyverse)  # For data manipulation and visualization
library(stargazer)  # For nice regression tables
require(gridExtra)
# If you haven't installed these packages yet, run:
# install.packages(c("foreign", "haven", "tidyverse", "stargazer"))
```

## Define the working directory and load data
```{r setwd}
# Set your working directory to where your data file is located
# Update the path below to match your file location
setwd("/Users/juhong/Desktop/S052/Data")
```

##Data: Thanks to Prof. Alejandro Ganimian and coauthors Karthik Muralidharan and Abhijeet Singh for this data
```{r importdata2}
# Load the data
dat <- read_dta("mindspark_data.dta")
origin <- read_dta("mindspark_data.dta")
```

## Q1: Exploratory Data Analysis
```{r explore}
# For this question, you will need:
# - The hist() function or ggplot() with geom_histogram() for histograms
# - The summarize() function from dplyr or the summary() function for descriptive statistics
# - You can calculate statistics by group using group_by() and summarize() in dplyr
# - Or use aggregate() or tapply() in base R
# 
# Hint: To get means by groups, you might use something like:
# dat %>%
   # group_by(variable1, variable2) %>%
   # summarize(mean_score = mean(math_std, na.rm = TRUE))

# Before turning to Google, AI, etc., try typing `?function_name` in the console to see documentation
# - Example: `?lm` for help with linear models
```

sum(is.na(dat))
dat <- drop_na(dat)

dat$treat %>% table()

head(dat)
413/(124+413)*100


counts_table <- table(dat$female)
print(prop.table(counts_table, margin = 1))


dat %>% group_by(female, treat) %>% summarize(mean_score = mean(math_std,na.rm = TRUE))

### Histogram of math scores on gender and treatment 
dat$female_f <- factor(dat$female,
                       levels = c(0, 1),
                       labels = c("Male", "Female"))

dat$treat_f <- factor(dat$treat,
                      levels = c(0, 1),
                      labels = c("Control", "Treatment"))
                      
gender_hist <- ggplot(dat, aes(x = math_std)) +
  geom_histogram(fill = "light blue",  #Set bar fill color, 
                 color = "black") + #Set bar border color
  facet_wrap(~female_f) 
  
treat_hist <- ggplot(dat, aes(x = math_std)) +
  geom_histogram(fill = "light pink", color = "black") + 
  facet_wrap(~treat_f) 
  
grid.arrange(grobs = list(gender_hist,treat_hist))


ggplot(dat, aes(x = female, y= math_std)) +
  geom_point(fill = "light blue", color = "black") 

### Histogram per group
f_t <- dat%>%filter(female == 1, treat == 1) %>% ggplot(aes(x=math_std))+geom_histogram()+ggtitle("Female in Treated")
f_c <- dat%>%filter(female == 1, treat == 0) %>% ggplot(aes(x=math_std))+geom_histogram()+ggtitle("Female in Controlled")
m_t <- dat%>%filter(female == 0, treat == 1) %>% ggplot(aes(x=math_std))+geom_histogram()+ggtitle("Male in Treated")
m_c <- dat%>%filter(female == 0, treat == 0) %>% ggplot(aes(x=math_std))+geom_histogram()+ggtitle("Male in Controlled")

grid.arrange(grobs = list(f_t, f_c, m_t, m_c))

### Histogram grouped by gender, colored by treat

library(dplyr)
library(ggplot2)
library(gridExtra)

m_all <- dat %>% 
  filter(female == 0) %>% 
  ggplot(aes(x = math_std,
             fill  = treat_f,
             group = treat_f)) +                        # group by treat
  geom_histogram(
    aes(y = after_stat(100 * count / sum(count))),     # % within each (male, treat)
    position = "identity",
    alpha = 0.6
  ) +
  ggtitle("Male") +
  labs(fill = "Group", y = "Percent for each group")

f_all <- dat %>% 
  filter(female == 1) %>% 
  ggplot(aes(x = math_std,
             fill  = treat_f,
             group = treat_f)) +
  geom_histogram(
    aes(y = after_stat(100 * count / sum(count))),     # % within each (female, treat)
    position = "identity",
    alpha = 0.6
  ) +
  ggtitle("Female") +
  labs(fill = "Group", y = "Percent for each group")

grid.arrange(grobs = list(f_all, m_all))


### Histogram of grouped by gender, colored by treat, counts percentaged 
### Histogram per group
f_t_dat <- dat%>%filter(female == 1, treat == 1) 
f_t <- f_t_dat %>%ggplot(aes(x=math_std, y= after_stat(100*count/ sum(count))))+geom_histogram()+ggtitle("Female in Treated")
f_c <- dat%>%filter(female == 1, treat == 0) %>% ggplot(aes(x=math_std))+geom_histogram()+ggtitle("Female in Controlled")
m_t <- dat%>%filter(female == 0, treat == 1) %>% ggplot(aes(x=math_std))+geom_histogram()+ggtitle("Male in Treated")
m_c <- dat%>%filter(female == 0, treat == 0) %>% ggplot(aes(x=math_std))+geom_histogram()+ggtitle("Male in Controlled")

grid.arrange(grobs = list(f_t, f_c, m_t, m_c))

## Q2: Separate and joint estimation of causal effects
```{r causal}
# Convert variables to factors if needed (this helps R handle categorical variables properly)
dat$treat <- as_factor(dat$treat)
dat$female <- as_factor(dat$female)

# For this question, you will need the lm() function for linear regression
# Remember to use robust standard errors!
m_all_dat <- dat %>%filter(female == 0)
# Model 1: Effect for male students only
# Hint: Use filter() or subset to select only male students (female == "0")
# lm(math_std ~ treat, data = ...)
head(m_all_dat)

mod1 <- lm(math_std ~ treat, data = m_all_dat)# Your code here
   
# If you want to see the full model output, including coefficient estimates, standard errors, test statistics, and p-values, call summary(mod).
summary(mod1)

# Model 2: Effect for female students only
f_all_dat <- dat%>%filter(female == 1)
mod2 <- lm(math_std ~ treat, data = f_all_dat)
summary(mod2)

# Model 3: Full model with interaction
# Hint: Use the formula: math_std ~ treat*female
# The * automatically includes main effects and interaction
mod3 <- lm(math_std ~ treat*female, data=dat)
summary(mod3)

# This code will output your regression results into a nice table
# (Similar to esttab in Stata)
stargazer(mod1, mod2, mod3, 
          type = "text",
          column.labels = c("Male", "Female", "Full"),
          title = "...WRITE A TITLE HERE...",
          star.char = c("*", "**", "***"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          notes = c("*p<0.05; **p<0.01; ***p<0.001"),
          notes.append = FALSE)


```
## Coding Pathway 

library(sjPlot)
library(sjmisc)
library(ggplot2)
theme_set(theme_sjplot())
head(dat)
plot_model(mod3, type="int")


## Additional Notes for Students:

### Helpful R Tips:
1. **Filtering data**: Use `filter()` from dplyr
   - Example: `filter(dat, female == 0)`

2. **Creating groups**: For analysis by groups, you can:
   - Use `group_by()` and `summarize()` from dplyr
   - Use the `by()` function in base R
   - Filter and analyze separately

3. **Getting help**: Type `?function_name` in the console to see documentation
   - Example: `?lm` for help with linear models

4. **Checking your work**: Use `summary()` on your model objects to see detailed output

5. **For the interaction model**: Remember that `treat*female` in the formula 
   automatically includes both main effects and the interaction term