---
title: |
  <center> S-052: Intermediate and Advanced Statistical Methods </center>
  <center> for Applied Educational Research </center>
  <center> Spring 2026 </center>
  <center> Week 3 Data Analytic Memo Starter Code </center>
author: "Sara Ji and Bailey Buchanan"
date: "Last Modified: Feb 2026"
output: pdf_document
---

### RQ: What is the causal effect of technology-led afterschool on math scores?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load in packages
```{r packages, warning = FALSE, message = FALSE}
# These packages will be helpful for your analysis
library(foreign)    # For reading various data formats
library(haven)      # For reading Stata files
library(tidyverse)  # For data manipulation and visualization
library(stargazer)  # For nice regression tables

# If you haven't installed these packages yet, run:
# install.packages(c("foreign", "haven", "tidyverse", "stargazer"))
```

## Define the working directory and load data
```{r setwd}
# Set your working directory to where your data file is located
# Update the path below to match your file location
#setwd("YOUR_DIRECTORY_PATH_HERE")
```

##Data: Thanks to Prof. Alejandro Ganimian and coauthors Karthik Muralidharan and Abhijeet Singh for this data
```{r importdata2}
# Load the data
dat <- read_dta("mindspark_data.dta")
head(dat)
```

## Q1: Exploratory Data Analysis
```{r explore}
# For this question, you will need:
# - The hist() function or ggplot() with geom_histogram() for histograms
# - The summarize() function from dplyr or the summary() function for descriptive statistics
# - You can calculate statistics by group using group_by() and summarize() in dplyr
# - Or use aggregate() or tapply() in base R
# 
# Hint: To get means by groups, you might use something like:
# dat %>%
   # group_by(variable1, variable2) %>%
   # summarize(mean_score = mean(math_std, na.rm = TRUE))

# Before turning to Google, AI, etc., try typing `?function_name` in the console to see documentation
# - Example: `?lm` for help with linear models

dat$treat <- factor(dat$treat,
                      levels = c(0 , 1),
                      labels = c("control", "treatment"))

dat$female <- factor(dat$female,
                      levels = c(0 , 1),
                      labels = c("male", "female"))

# question one: a) What percentage of students in the sample were female?
percent_female <- mean(dat$female == "female") * 100
# 75.92892 %
percent_male <- mean(dat$female == "male") * 100

total_gap <- percent_female - percent_male

# participation gap is 75.92892 - 24.07108 = 51.85784 percentage points

# questions
# b) What was the gender gap in endline scores among students who were not offered treatment?
# c) What was the gender gap in endline scores among students who were offered treatment?

gender_by_treat <- dat |>
  group_by(treat) |>
  summarise(
    pct_female = mean(female == "female") * 100 ,
    pct_male = mean(female == "male") * 100,
    gap = pct_female - pct_male
  ) 

?group_by


# question
# d) Plot the distribution of endline scores for each of the four groups we consider in this
# analysis (female students offered treatment, female students not offered treatment,
# male students offered treatment, male students not offered treatment). Based on these
# four distributions, what do you hypothesize about (1) the treatment effect for male
# students, (2) the treatment effect for female students, and (3) whether there is a
# difference between the effects? Explain in 3-5 sentences, making reference to your
# histograms.


#### fix the percent axis!!! 

ggplot(dat, aes(x = female, fill = treat)) +
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Distribution of treatment and control groups by gender",
    x = "Gender",
    y = "Percent",
    fill = "Group"
  ) +
  theme_minimal()


?stat=count

###### questions answered up to this point. Below it's just some code to help us get started with the next questions ######

ggplot(dat, aes(x = math_std)) +
  geom_histogram(binwidth = .1) +
  labs(title = "Distribution of Standardized Math Scores",
       x = "Standardized Math Score",
       y = "Frequency")

reg1 <- lm(math_std ~ treat, data = dat) 
summary(reg1) 


# ggplot(dat, aes(x = st_id, y = math_std)) +
#   geom_smooth(method = "lm") +
#   labs(title = "Math Scores by Treatment Status",
#        x = "Treatment Status",
#        y = "Standardized Math Score")

head(dat)
?factor
```

## Q2: Separate and joint estimation of causal effects
```{r causal}
# Convert variables to factors if needed (this helps R handle categorical variables properly)
dat$treat <- as_factor(dat$treat)
dat$female <- as_factor(dat$female)

# For this question, you will need the lm() function for linear regression
# Remember to use robust standard errors!

# Model 1: Effect for male students only
# Hint: Use filter() or subset to select only male students (female == "0")
# lm(math_std ~ treat, data = ...)
mod1 <- lm(math_std ~ treat, data = filter(dat, female == "male"), na.action = "na.exclude")
summary(mod1)
   
# If you want to see the full model output, including coefficient estimates, standard errors, test statistics, and p-values, call summary(mod).

# Model 2: Effect for female students only  
mod2 <- lm(math_std ~ treat, data = filter(dat, female == "female"), na.action = "na.exclude")
summary(mod2)

# Model 3: Full model with interaction
# Hint: Use the formula: math_std ~ treat*female
# The * automatically includes main effects and interaction
mod3 <- lm(math_std ~ treat*female, data = dat, na.action = "na.exclude")
summary(mod3)

# This code will output your regression results into a nice table
# (Similar to esttab in Stata)
stargazer(mod1, mod2, mod3, 
          type = "text",
          column.labels = c("Male", "Female", "Full"),
          title = "...WRITE A TITLE HERE...",
          star.char = c("*", "**", "***"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          notes = c("*p<0.05; **p<0.01; ***p<0.001"),
          notes.append = FALSE)


```


## Additional Notes for Students:

### Helpful R Tips:
1. **Filtering data**: Use `filter()` from dplyr
   - Example: `filter(dat, female == 0)`

2. **Creating groups**: For analysis by groups, you can:
   - Use `group_by()` and `summarize()` from dplyr
   - Use the `by()` function in base R
   - Filter and analyze separately

3. **Getting help**: Type `?function_name` in the console to see documentation
   - Example: `?lm` for help with linear models

4. **Checking your work**: Use `summary()` on your model objects to see detailed output

5. **For the interaction model**: Remember that `treat*female` in the formula 
   automatically includes both main effects and the interaction term