---
title: "Week 8 DAM Starter Code"
date: "2025-10-15"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Set working directory 
```{r setwd}
#Set working directory. Uncomment the next line of code and run it. 
#setwd("[YOUR FILE PATH HERE]")

#example: setwd("~/S52 2025/Misc")
```

## Load Packages 
```{r packages}
library(haven)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(stargazer)
library(lme4)
library(psych)
```

## Load Data 
```{r data}
medicaldf <- read_dta("KoreanMedicalPortfolios.dta")
sedadf <- read_dta("SEDA19.dta")
```

##Step 1: Exploratory Data Analysis of SEDA SES relationships
```{r explore}
#Drop DC and HI which each have only have 1 school district.
sedadf <- sedadf %>%
  filter(stateabb != "DC", stateabb != "HI")

#Set plot range data
sedadf$plotrange <- (sedadf$sesavgall < 3)  & (sedadf$sesavgall > -4) & (sedadf$meanavg > 1) & (sedadf$meanavg < 10)

#Trellis plot illustrating separate regression lines by state
sedadf %>%
  filter(plotrange == TRUE) %>% #plot only data that is in the range set above
  ggplot(aes(x = sesavgall, y = meanavg)) + 
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE) + #plot an lm line 
  facet_wrap(~stateabb) + #plot each state separately
  xlab("District Socioeconomic Status") +
  ylab("District Average Test Score, 2009-2016") +
  ggtitle("trellis")

#Calculating intercepts and slopes for each state
sedadf <- sedadf %>%
  group_by(stateabb) %>%
  mutate(intercept = coef(lm(meanavg ~ sesavgall))[1],
         slope = coef(lm(meanavg ~ sesavgall))[2])

#Optional example to open "object" > coefficients > see the 1st and 2nd element (to see where the above estimates are stored)
#object <- lm(meanavg ~ sesavgall, data = dat)
  

#Predicted average test scores for each district given SES
sedadf$predicted <- sedadf$intercept + sedadf$slope*sedadf$sesavgall

#Overlaid linear fit of regression lines for all 49 states
ggplot(sedadf, aes(x = sesavgall, y = predicted, group = stateabb)) + 
  geom_smooth() + 
  scale_y_continuous(breaks = 1:9) +
  xlab("District Socioeconomic Status") +
  ylab("District Average Test Score, 2009-2016") +
  ggtitle("overlaid")
```

##Step 2: Building multilevel regression models
```{r mlm}
#Establish a binary variable for states that had growing public school enrollments from 2009 to 2016.
#Data from https://nces.ed.gov/programs/digest/d19/tables/dt19_203.20.asp

# enroll <- readxl::read_excel("tabn203.20.xls")
# names(enroll) <- as.character(enroll[2, ])
# enroll <- enroll[-c(1:10), ]
# enroll <- enroll[-c(52:66), ]
# 
# names(enroll)
# 
# names(enroll) <- sub("\\\\.*", "", names(enroll))
# 
# enroll$`Fall 2009` <- as.numeric(enroll$`Fall 2009`)
# enroll$`Fall 2016` <- as.numeric(enroll$`Fall 2016`)
# 
# enroll$growth <- as.integer(enroll$`Fall 2016` > enroll$`Fall 2009`)
# 
# keep_states <- c(
#   "Alabama", "Connecticut", "Illinois", "Maine", "Michigan",
#   "Mississippi", "Missouri", "New Hampshire", "New York",
#   "Ohio", "Pennsylvania", "Rhode Island", "Vermont",
#   "West Virginia", "Wisconsin"
# )
# 
# enroll <- enroll[enroll[[1]] %in% keep_states, ]

#Step 2A: Generate an indicator variable here that equals 1 for the following states:
		#AL, CT, IL, ME, MI, MS, MO, NH, NY, OH, PA, RI, VT, WV, WI


growing_states <- c("AL","CT","IL","ME","MI","MS","MO","NH","NY","OH","PA","RI","VT","WV","WI")

sedadf$growing_states <- ifelse(sedadf$stateabb %in% growing_states, 1, 0)


#Model 0: Random Intercepts baseline model to establish a baseline log-likelihood

#Run your intercept-only model, using the lmer command (make sure REML = FALSE to use maximum likelihood modeling)
#Important: Restrict your regression to districts with non-missing SES so that the sample is the same between this regression and the following regressions.

sedadfmod <- sedadf %>%
  filter(!is.na(sesavgall))

mod0 <- lmer(
  meanavg ~ 1 + (1 | stateabb),
  data = sedadfmod,
  REML = FALSE
)
summary(mod0)
#Model 1: Fixed intercepts model to fit within-state slopes
#This code runs a fixed effects model. We provide it since it is not a focus of this assignment.

mod1 <- lm(meanavg ~ sesavgall + stateabb, data = sedadfmod)
summary(mod1)

#Model 2: Random intercepts model with SES (level-1 predictor)
#Here, run a random intercepts regression of test scores on SES, again using lmer and REML=FALSE 

mod2 <- lmer(
  meanavg ~ sesavgall + (1 | stateabb),
  data = sedadfmod,
  REML = FALSE
)
summary(mod2)

#Model 3: Random intercepts model with level-2 predictor
#Now add the level-2 predictor "growing" to your regression.

mod3 <- lmer(
  meanavg ~ sesavgall + growing_states + (1 | stateabb),
  data = sedadfmod,
  REML = FALSE
)
summary(mod3)


#note neg2ll (deviance) doesn't come out of mod1, so we have to craft the input a bit
neg2ll <- round(as.numeric(unlist(lapply(list(mod0, mod2, mod3), FUN = function(x) getME(x, "devcomp")$cmp['dev']))), 2)
neg2ll <- c("", neg2ll)
neg2ll <- neg2ll[c(2, 1, 3, 4)]


#same with sigma2_u (between group variance) and sigma2_e (within group variance)
temp <- do.call(rbind, lapply(list(mod0, mod2, mod3), FUN = function(x) data.frame(VarCorr(x))))
  #store between group variance 
sigma2_u <- round(temp$vcov[c(1, 3, 5)], 3)
sigma2_u <- c("", sigma2_u)
sigma2_u <- sigma2_u[c(2, 1, 3, 4)]
  #store within group variance 
sigma2_e <- round(temp$vcov[c(2, 4, 6)], 3)
sigma2_e <- c("", sigma2_e)
sigma2_e <- sigma2_e[c(2, 1, 3, 4)]

#Generate model building tables
stargazer(mod0, mod1, mod2, mod3, 
          type = "text",
          column.labels = paste("Model", 0:3),
          model.numbers = FALSE,
          keep = c("sesavgall", "growing", "Constant"), #specify the variables to include in the table and avoid printing all Mod1 fixed effects
          keep.stat = c("n"), #keep n (number of observations)
          add.lines = list(c("neg2ll", neg2ll),
                           c("sigma2u", sigma2_u),
                           c("sigma2e", sigma2_e)))


#Model 4: Fixed effects model with level 2 predictor. Why does this fail?	
mod4 <- lm(meanavg ~ stateabb + sesavgall + growing_states, data = sedadfmod)
summary(mod4)

sedadfmod %>%
  group_by(stateabb) %>%
  summarise(
    n = n(),
    var_grow = var(growing_states, na.rm = TRUE)
  )
```

## Step 3: Supplementary visualization: A caterpillar plot to visualize random intercepts
```{r supp}
#Fit a mixed model equivalent to Model 3 and save random intercepts
summary(mod2)
temp <- ranef(mod2, condVar = TRUE)
randeffs <- data.frame(state = row.names(temp$stateabb),
                       int = as.numeric(unlist(temp$stateabb)),
                       stderr = as.numeric(sqrt(data.frame(attr(temp$stateabb, "postVar")))))
randeffs$rank <- rank(randeffs$int)

ggplot(randeffs, aes(x = rank, y = int)) +
  geom_text(aes(label = state, y = int + 2*stderr), nudge_y = 0.07, angle = 90) +
  geom_errorbar(aes(ymax = int + 2*stderr, ymin = int - 2*stderr)) +
  geom_point(size = 2) +
  geom_hline(aes(yintercept = 0), col = "red") +
  xlab("Rank") +
  ylab("Random state intercept with 95% CI, in grade levels") +
  ggtitle("caterpillar")
```

# Part 2: Interrater and interitem reliability

##RQ1 Data: Intraclass Correlations for Interrater Reliability (Raters Nested)
```{r rq1}
#histogram of all scores
ggplot(medicaldf, aes(x = score)) + 
  geom_histogram(binwidth = 1, color = "lightblue") + 
  ggtitle("allscores")

#boxplot of rater scores
ggplot(medicaldf, aes(group = rater, x = rater, y = score)) + 
  geom_boxplot() + ggtitle("raterscores")

#tibble of rater scores (mean, standard deviation, frequency)
medicaldf %>%
  group_by(rater) %>%
  summarize(scoremean = mean(score),
            scoresd = sd(score),
            freq = n())

#boxplot of item scores
ggplot(medicaldf, aes(group = item, x = rater, y = score)) + 
  geom_boxplot() + 
  ggtitle("itemscores")

#tibble of item scores (mean, standard deviation, frequency)
medicaldf %>%
  group_by(item) %>%
  summarize(scoremean = mean(score),
            scoresd = sd(score),
            freq = n())

#Begin with the overall score:

#Estimate intraclass correlations for each item (create a loop)
for(i in 1:5){ #the loop iterates from 1 to 5 
  assign(paste0("moditem", i), #create an object for each
         lmer(score ~ (1|student), data = filter(medicaldf, item == i))) #run a random intercept model for each unique student, the data used for each model will be a subset where the item is the current value/iteration of "i" (1-5)
}
#^result is 5 models stored in environment


temp <- do.call(rbind, lapply(list(moditem1, moditem2, moditem3, moditem4, moditem5), FUN = function(x) data.frame(VarCorr(x))))
sigma2_u <- round(temp$vcov[c(1, 3, 5, 7, 9)], 3)
sigma2_e <- round(temp$vcov[c(2, 4, 6, 8, 10)], 3)
rho <- round(sigma2_u / (sigma2_u + sigma2_e), 3) #ICC (inter-rater reliability)

stargazer(moditem1, moditem2, moditem3, moditem4, moditem5, type = "text",
          title = "Interrater reliability coefficients for item scores",
          column.labels = c("Goals", "Processes", "Reflect", "Plans", "Clarity"),
          model.numbers = FALSE,
          keep.stat = c("n"),
          add.lines = list(c("sigma2u", sigma2_u),
                           c("sigma2e", sigma2_e),
                           c("rho", rho)))

#OPTIONAL:
  #Estimate the intraclass correlation for the average of the 5 item scores
  medicaldf <- medicaldf %>%
    group_by(student, rater) %>%
    mutate(avgscore = mean(score))

  #Estimate intraclass correlation for each rater's average score
  lmer(avgscore ~ (1|student), data = filter(medicaldf, item == 1))
```

## RQ2 Analysis: Intraclass Correlations for Interitem Reliability (Items Crossed)
```{r rq2}
#Estimate intraclass correlations for each rater
for(i in 1:5){
  assign(paste0("modrater", i),
         lmer(score ~ (1|student) + (1|item), data = filter(medicaldf, rater == i)))
}


temp <- do.call(rbind, lapply(list(modrater1, modrater2, modrater3, modrater4, modrater5), FUN = function(x) data.frame(VarCorr(x))))
sig2pers <- round(temp$vcov[c(1, 4, 7, 10, 13)], 4) #person var (good)
sig2item <- round(temp$vcov[c(2, 5, 8, 11, 14)], 4) #item var (neutral)
sig2e <- round(temp$vcov[c(3, 6, 9, 12, 15)], 4) #error var (bad)

stargazer(modrater1, modrater2, modrater3, modrater4, modrater5, type = "text",
          title = "Interitem reliability coefficients by rater",
          column.labels = paste("Rater", 1:5, sep = " "),
          model.numbers = FALSE,
          keep.stat = c("n"),
          add.lines = list(c("sig2pers", sig2pers),
                           c("sig2item", sig2item),
                           c("sig2e", sig2e)))
```

###Cronbach's alpha for item scores for Rater 1
```{r rq3}
#Focus only on the first rater for alpha (transform data from long to wide to prep for alpha command)
dat_r1 <- medicaldf %>%
  filter(rater == 1) %>%
  ungroup() %>%
  dplyr::select(student,item,score) %>%
  pivot_wider(names_from = item, values_from = score, names_prefix = "item")



#Cronbach's alpha
statobj <- alpha(dplyr::select(dat_r1, -student))
statobj$total$raw_alpha #unstandardized cronbach's alpha

statobj$item.stats %>% dplyr::select(raw.r) #item-test correlations
```

###Coding-Pathway
```{r coding}
# 6) Let’s merge and explore these files to see where our current dataset is coming from!
# This may require some patience and searching on your part, so try to get an early start on
# this. 
seda_pool <- read_dta("SEDA_cov_geodist_pool_v30(1).dta")
seda_gcs <- read_dta("SEDA_geodist_pool_gcs_v30(1).dta")

# only merged items from both datasets
inner_join <- inner_join(seda_gcs, seda_pool, by = "leaid")

class(seda_gcs$leaidC)
names(seda_gcs)

class(seda_pool$leaid)

seda_gcs$leaid <- as.numeric(seda_gcs$leaidC)
dim(seda_gcs)
dim(seda_pool)

seda_gcs <- seda_gcs %>%
  mutate(fips = as.integer(fips))

seda_pool <- seda_pool %>%
  mutate(fips = as.integer(fips))

leftseda_merged <- left_join(seda_gcs, seda_pool, by = "leaid")
names(leftseda_merged)


# c)	Calculate the mean of the mn\_avg\_eb variable (familiar to us as meanavg)
# for the “all” subgroup. Confirm that there are 13,002 observations with a mean 
# of 5.561088. Paste your output below. 

means <- leftseda_merged %>%
  filter(!is.na(leaid)) %>% 
  # filter(!is.na(mn_avg_eb)) %>%  
  filter(subgroup == "all") %>%
  summarise(
    n_obs = n(),
    mean_meanavg = mean(mn_avg_eb, na.rm= TRUE)
  )

print(means)

# d) As we did in Week 1, regress mn\_avg\_eb on sesavgall for the “all” 
# subgroup. Confirm the regression coefficient is .9084559 with 12,601 
# observations. Paste your output below.

df <- leftseda_merged %>%
  filter(!is.na(leaid)) %>%
  filter(!is.na(mn_avg_eb)) %>%
  filter(!is.na(sesavgall)) %>%
  filter(subgroup == "all") 


mod01 <- lm(mn_avg_eb ~ sesavgall, data = df )

summary(mod01)


# e)	Explain why there are more districts in part (c) than in the regression 
# in part (d) (i.e., what information is missing for the 401 districts that 
# appear in part (c), but not part (d)?). List a few of the “missing” districts 
# here. (Optional: do you notice any common characteristics of these districts?)


df_missing <- leftseda_merged %>%
  filter(!is.na(leaid)) %>%
  filter(!is.na(mn_avg_eb)) %>%
  filter(is.na(sesavgall)) %>%
  filter(subgroup == "all")

print(df$leanm)

```
